---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---
```{r setoptions,echo=FALSE}
# opt_chunk$set(echo = FALSE, results = "hide")

```

## Loading and preprocessing the data
1. Loading Data  
```{r LoadingData, echo=TRUE, message=FALSE, warning=FALSE}
# Loading libraries 
library(data.table)
library(dplyr)
library(ggplot2)
library(lubridate)
library(plyr)

# Loading data
df <- fread("activity.csv")
head(df)
```

2. **Preprocessing data**:  a new variable "timeInterval" is created, containing the conversion of the variable "interval" from integer to POSIXct although this had no use for us later.  
```{r preprocessing, echo=TRUE, message=FALSE, warning=FALSE}
to_time <- function(x){
    # returns the 5-minutes interval in "hh:nn" format.
    sprintf("%02d:%02d", x%/%100, x%%100)
}
df$timeInterval <- with(df, as.POSIXct(paste(date, to_time(interval)), format="%Y-%m-%d %H:%M"))
head(df)
```
## What is mean total number of steps taken per day?  
1. Calculating the total number of steps taken each day  
```{r totalstepsNumber, echo=TRUE, message=FALSE, warning=FALSE}
totalStepsPerDay <- df %>% dplyr::group_by(date) %>% dplyr::summarise(totalSteps = sum(steps, na.rm = TRUE))
head(totalStepsPerDay)
```

2. Histogram of the total number of steps taken each day  

```{r histogram, echo=TRUE, message=FALSE, warning=FALSE, fig.height=5, fig.width=6}
qplot(totalStepsPerDay$totalSteps, binwidth = 1000, xlab = "Total number of steps per day", ylab = "Frequency", main = "Total number of steps taken each day", fill = "red") + guides(fill = FALSE)
```  
  
3. Calculating the mean and median of the total number of steps taken per day    

```{r meanAndMedian, echo=TRUE, message=FALSE, warning=FALSE, fig.height=5, fig.width=6}
meanTotalStepsPerDay <- mean(totalStepsPerDay$totalSteps, na.rm = TRUE)
medianTotalStepsPerDay <- median(totalStepsPerDay$totalSteps, na.rm = TRUE)
```
- The mean of the total number of steps taken per day is: **`r meanTotalStepsPerDay`**  
- The median of the total number of steps taken per day is: **`r medianTotalStepsPerDay`**  

## What is the average daily activity pattern?
1.  The average number of steps taken across all days  
```{r plot, echo=TRUE, message=FALSE, warning=FALSE, fig.height=5, fig.width=6}
dailyAverage <- df %>% dplyr::group_by(interval) %>% dplyr::summarise(dailyAverageSteps = mean(steps, na.rm = TRUE))
# plotting
qplot(interval, dailyAverageSteps, data = dailyAverage, geom = "line", col = "red", xlab = "5-minute interval", ylab = "The average number of steps taken", main = "Average daily activity") + guides(col = FALSE)
```

2. The 5-minute interval containing the maximum number of steps
```{r maxStepsInterval, echo=TRUE, message=FALSE, warning=FALSE}
maxInterval <- dailyAverage %>% filter(dailyAverageSteps == max(dailyAverageSteps))
```
- The 5-minute interval containing the maximum number of steps is **`r to_time(maxInterval$interval)`**  

## Imputing missing values
1. Calculating the total number of missing values in the dataset  

```{r numberNAs, echo=TRUE, message=FALSE, warning=FALSE}
nbrNA <- sum(!complete.cases(df))
```
- The total number of missing values is: **`r nbrNA`**  

2. Missing values can be treated using several methods such as **mean/mode/median** imputation, **KNN** imputation or using a **prediction model** to estimate values that will substitute the missing data.   
In the following an implementation of an imputation method based on the mean.  
```{r imputeNAsMean, echo=TRUE, message=FALSE, warning=FALSE}
imputeByMean <- function(x) replace(x, is.na(x), mean(x, na.rm = TRUE))
```
3. Creation of a new dataset with the missing data filled in by the mean for 5-minute interval.   
```{r dfWithoutNAs, echo=TRUE, message=FALSE, warning=FALSE}
# Apply mean imputation after grouping data by interval
df2 <- df %>% dplyr::group_by(interval) %>% dplyr::mutate(steps = imputeByMean(steps))
head(df2)
```
4. Histogram of the total number of steps taken each day  

```{r totalstepsNumber2, echo=TRUE, message=FALSE, warning=FALSE, fig.height=5, fig.width=6}
totalStepsPerDay2 <- df2 %>% dplyr::group_by(date) %>% dplyr::summarise(totalSteps = sum(steps, na.rm = TRUE))
# histogram
qplot(totalStepsPerDay2$totalSteps, binwidth = 1000, xlab = "Total number of steps per day", ylab = "Frequency", main = "New total number of steps taken each day", fill = "red") + guides(fill = FALSE)
# Mean & Median
meanTotalStepsPerDay2 <- mean(totalStepsPerDay2$totalSteps)
medianTotalStepsPerDay2 <- median(totalStepsPerDay2$totalSteps)
```
- The new mean of the total number of steps taken per day is: **`r  meanTotalStepsPerDay2`**  
- The new median of the total number of steps taken per day is: **`r medianTotalStepsPerDay2`**  
- The new values of the mean, as well as the median, are visibly larger. This is because many of the null values have been replaced with the 5-minute interval averages after applying the imputation.  

## Are there differences in activity patterns between weekdays and weekends?  

1. Create a new factor variable  indicating whether a given date is a weekday or weekend day.    

```{r factors, echo=TRUE, message=FALSE, warning=FALSE}
df2 <- df2 %>% dplyr::mutate(weekDay = factor(1*(weekdays(date) %in% c("Saturday", "Sunday")), labels = c("weekday", "weekend") ))
head(df2)
```

2. The average number of steps taken across all weekday days or weekend days

```{r plot2, echo=TRUE, message=FALSE, warning=FALSE, fig.height=5, fig.width=6}
weekdayAverage <- df2 %>% dplyr::group_by(interval, weekDay) %>% dplyr::summarise(weekdayAverageSteps = mean(steps, na.rm = TRUE))
# plotting
qplot(interval, weekdayAverageSteps, data = weekdayAverage, geom = "line", col = "red", facets = weekDay ~ ., xlab = "5-minute interval", ylab = "The average number of steps", main = "Average weekday/weekend activity") + guides(col = FALSE)
```



